---
title: "COVID-19 Data Analysis"
author: ""
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

# Setup and Data Loading

```{r setup, include=TRUE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Load required libraries
library(tidyverse)
library(lubridate)
library(scales)
library(viridis)
library(leaflet)
library(kableExtra)

# Load data from JHU CSSE GitHub repository
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c(
  "time_series_covid19_deaths_global.csv",
  "time_series_covid19_confirmed_global.csv",
  "time_series_covid19_confirmed_US.csv",
  "time_series_covid19_deaths_US.csv"
)
urls <- str_c(url_in, file_names)

# Read the data
global_deaths <- read.csv(urls[1])
global_cases <- read.csv(urls[2])
us_cases <- read.csv(urls[3])
us_deaths <- read.csv(urls[4])
```

# Required libraries

tidyverse

lubridate

scales

viridis

leaflet

kableExtra

# Examine Data and get a random sample

```{r examine_data, include=TRUE}

convert_to_long <- function(df, value_type = NULL) {
  if (is.null(value_type)) {
    value_type <- if ("deaths" %in% names(df)) "deaths" else "cases"
  }
  
  df %>%
    pivot_longer(cols = starts_with("X"),
                 names_to = "date",
                 values_to = value_type) %>%
    mutate(date = as.Date(gsub("X","",date), format="%m.%d.%y"))
}

# Then use it like this:
global_cases_long <- convert_to_long(global_cases, "cases")
global_deaths_long <- convert_to_long(global_deaths, "deaths")
us_cases_long <- convert_to_long(us_cases, "cases")
us_deaths_long <- convert_to_long(us_deaths, "deaths")

# Get a random sample of data

sample_and_display <- function(data, n, caption) {
  data %>%
    slice_sample(n = n) %>%
    kable("pipe", caption = caption) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}

# Use the function:
sample_and_display(global_deaths_long, 5, "Random Sample of Global Deaths Data")
sample_and_display(global_cases_long, 5, "Random Sample of Global Cases Data")
sample_and_display(us_deaths_long, 5, "Random Sample of US Deaths Data")
sample_and_display(us_cases_long, 5, "Random Sample of US Cases Data")

```

# Data Overview

All datasets have columns for location information:
Province.State / Province_State: Represents states and provinces.
Country.Region / Country_Region: Represents the country.
Lat and Long / Long_: Represent the latitude and longitude of the location.

All datasets have date columns (e.g., X1.22.20, X1.23.20), which represent cumulative data for each day starting from January 22, 2020.

The datasets are wide-format time series data, where each row corresponds to a specific location, and the columns after the location information represent cumulative counts for each day.

global_cases and global_deaths contain worldwide data, with rows for each country.
us_cases and us_deaths focus specifically on the United States with some additional information such as:
Columns like Admin2 (county-level data), FIPS (Federal Information Processing Standards code), and Combined_Key (a unique identifier for each location).
us_deaths also includes a Population column, which is not present in the other datasets.

global_cases and us_cases track confirmed cases.
global_deaths and us_deaths track deaths.

global_cases and global_deaths are directly related:
They share the same structure and location columns (Province.State, Country.Region, Lat, Long).
The date columns in both datasets represent cumulative counts of cases and deaths, respectively, for the same locations and dates.
These datasets can be joined on Province.State, Country.Region, Lat, and Long to analyze the relationship between cases and deaths globally.

us_cases and us_deaths are also directly related:
They share the same structure and location columns (UID, FIPS, Admin2, Province_State, Country_Region, Lat, Long_, Combined_Key).
The date columns in both datasets represent cumulative counts of cases and deaths, respectively, for the same locations and dates.
These datasets can be joined on UID or Combined_Key to analyze the relationship between cases and deaths in the US.

# Checking for any missing data

```{r missing_data, include=TRUE}

# Function to create comprehensive data quality summary
create_comprehensive_summary <- function(data, dataset_name) {
  complete_cases <- sum(complete.cases(data))
  incomplete_cases <- nrow(data) - complete_cases
  
  data.frame(
    Dataset = dataset_name,
    Total_Rows = nrow(data),
    Complete_Rows = complete_cases,
    Incomplete_Rows = incomplete_cases,
    Pct_Complete = round(complete_cases / nrow(data) * 100, 1),
    Pct_Incomplete = round(incomplete_cases / nrow(data) * 100, 1),
    Total_Columns = ncol(data),
    Columns_Missing = sum(colSums(is.na(data)) > 0)
  )
}

# Apply to all datasets
comprehensive_summary <- bind_rows(
  create_comprehensive_summary(global_cases, "Global Cases"),
  create_comprehensive_summary(global_deaths, "Global Deaths"),
  create_comprehensive_summary(us_cases, "US Cases"),
  create_comprehensive_summary(us_deaths, "US Deaths")
)

# Display formatted table
comprehensive_summary %>%
  kable("pipe", 
        col.names = c("Dataset", "Total Rows", "Complete Rows", 
                     "Incomplete Rows", "% Complete", "% Incomplete",
                     "Total Columns", "Columns with Missing"),
        caption = "Data Quality Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

# Data Preprocessing

```{r preprocessing_data, include=TRUE}

# Data Preprocessing

# Convert wide format to long format
convert_to_long <- function(df, value_type = NULL) {
  if (is.null(value_type)) {
    value_type <- if ("deaths" %in% names(df)) "deaths" else "cases"
  }
  
  df %>%
    pivot_longer(cols = starts_with("X"),
                 names_to = "date",
                 values_to = value_type) %>%
    mutate(date = as.Date(gsub("X","",date), format="%m.%d.%y"))
}

# Apply long format conversion
global_cases_long <- convert_to_long(global_cases, "cases")
global_deaths_long <- convert_to_long(global_deaths, "deaths")
us_cases_long <- convert_to_long(us_cases, "cases")
us_deaths_long <- convert_to_long(us_deaths, "deaths")

# Create global combined dataset
global_data <- global_cases_long %>%
  full_join(global_deaths_long) %>%
  group_by(date) %>%
  summarize(total_cases = sum(cases),
            total_deaths = sum(deaths))

# Calculate daily cases and identify top countries
daily_cases_by_country <- global_cases_long %>%
  arrange(Country.Region, date) %>%
  group_by(Country.Region) %>%
  mutate(new_cases = cases - lag(cases, default = first(cases))) %>%
  ungroup()

# Get top 10 and top 5 countries
top_10_countries <- daily_cases_by_country %>%
  group_by(Country.Region) %>%
  summarize(total_cases = max(cases)) %>%
  arrange(desc(total_cases)) %>%
  head(10) %>%
  pull(Country.Region)

top_5_countries <- daily_cases_by_country %>%
  group_by(Country.Region) %>%
  summarize(total_cases = max(cases)) %>%
  arrange(desc(total_cases)) %>%
  head(5) %>%
  pull(Country.Region)

```

# Create summary table

```{r global_summary, include=TRUE}

global_summary <- global_data %>%
  summarize(
    total_cases = max(total_cases),
    total_deaths = max(total_deaths),
    overall_cfr = (total_deaths/total_cases)*100
  )

global_summary %>%
  gather(metric, value) %>%
  mutate(
    value = ifelse(metric == "overall_cfr",
                  sprintf("%.2f%%", value),
                  formatC(value, format="d", big.mark=","))
  ) %>%
  kable("pipe", col.names = c("Metric", "Value"),
        caption = "Global COVID-19 Summary Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


# Visualizations

### Global cumulative cases and deaths over time

```{r visualize_cummulative, include=TRUE}

# Global cumulative cases and deaths over time
ggplot(global_data, aes(x=date)) +
  geom_line(aes(y=total_cases, color="Cases"), size=1) +
  geom_line(aes(y=total_deaths, color="Deaths"), size=1) +
  scale_y_log10(labels = scales::comma) +
  scale_color_manual(values=c("Cases"="blue", "Deaths"="red")) +
  labs(title="Global COVID-19 Cases and Deaths (Log Scale)",
       x="Date",
       y="Count (log scale)",
       color="Type") +
  theme_minimal() +
  theme(legend.position="bottom")
```

### Top 10 countries by total cases

```{r visualize_top_10_countries_cases, include=TRUE}

# Top 10 countries by total cases
top_10_countries <- global_cases_long %>%
  group_by(Country.Region) %>%
  summarize(total_cases = max(cases)) %>%
  arrange(desc(total_cases)) %>%
  head(10)

ggplot(top_10_countries, aes(x=reorder(Country.Region, total_cases), y=total_cases)) +
  geom_bar(stat="identity", fill="steelblue") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(title="Top 10 Countries by Total COVID-19 Cases",
       x="Country",
       y="Total Cases") +
  theme_minimal()
```

### Top 10 US states by total cases

```{r top_10_us_cases, include=TRUE}


# US specific analysis
us_cases_long <- us_cases %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "date",
               values_to = "cases") %>%
  mutate(date = as.Date(gsub("X","",date), format="%m.%d.%y"))

# Top 10 US states by total cases
top_10_states <- us_cases_long %>%
  group_by(Province_State) %>%
  summarize(total_cases = max(cases)) %>%
  arrange(desc(total_cases)) %>%
  head(10)

ggplot(top_10_states, aes(x=reorder(Province_State, total_cases), y=total_cases)) +
  geom_bar(stat="identity", fill="darkgreen") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(title="Top 10 US States by Total COVID-19 Cases",
       x="State",
       y="Total Cases") +
  theme_minimal()

```

### COVID-19 Case Fatality Rate by Country and US State

```{r fatality_by_country, include=TRUE}

# Case Fatality Rate by Country
cfr_by_country <- global_cases_long %>%
  inner_join(global_deaths_long, by = c("Country.Region", "date")) %>%
  group_by(Country.Region) %>%
  summarize(total_cases = max(cases),
            total_deaths = max(deaths),
            cfr = (total_deaths/total_cases)*100) %>%
  filter(total_cases > 10000) %>%  # Filter for countries with significant cases
  arrange(desc(cfr)) %>%
  head(10)

# Case Fatality Rate by US State
cfr_by_state <- us_cases_long %>%
  filter(date == max(date)) %>%  # Filter for latest date first
  group_by(Province_State) %>%
  summarize(total_cases = sum(cases)) %>%
  left_join(
    us_deaths_long %>%
      filter(date == max(date)) %>%
      group_by(Province_State) %>%
      summarize(total_deaths = sum(deaths)),
    by = "Province_State"
  ) %>%
  mutate(cfr = (total_deaths/total_cases)*100) %>%
  filter(total_cases > 10000) %>%  # Filter for US States with significant cases
  arrange(desc(cfr)) %>%
  head(10)

# Case Fatality Rate by Country
ggplot(cfr_by_country, 
       aes(x = reorder(Country.Region, cfr), y = cfr)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() +
  labs(title = "COVID-19 Case Fatality Rate by Country",
       subtitle = "Top 10 countries with at least 10,000 cases",
       x = "Country",
       y = "Case Fatality Rate (%)") +
  theme_minimal()

# Case Fatality Rate by State
ggplot(cfr_by_state, 
       aes(x = reorder(Province_State, cfr), y = cfr)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() +
  labs(title = "COVID-19 Case Fatality Rate by US State",
       subtitle = "Top 10 states with at least 10,000 cases",
       x = "State",
       y = "Case Fatality Rate (%)") +
  theme_minimal()

```

### Global analysis with population adjustment 

US population data is included in the data set. 

I am adding global population from an external source.

```{r global_analysis_adjusted, include=TRUE}

# Create population data for countries (2023 estimates)
population_data_countries <- data.frame(
  Country.Region = c(
    "Afghanistan", "Albania", "Algeria", "Andorra", "Angola", "Antigua and Barbuda", "Argentina", "Armenia", "Australia", "Austria",
    "Azerbaijan", "Bahamas", "Bahrain", "Bangladesh", "Barbados", "Belarus", "Belgium", "Belize", "Benin", "Bhutan",
    "Bolivia", "Bosnia and Herzegovina", "Botswana", "Brazil", "Brunei", "Bulgaria", "Burkina Faso", "Burma", "Burundi", "Cabo Verde",
    "Cambodia", "Cameroon", "Canada", "Central African Republic", "Chad", "Chile", "China", "Colombia", "Comoros", "Congo (Brazzaville)",
    "Congo (Kinshasa)", "Costa Rica", "Cote d'Ivoire", "Croatia", "Cuba", "Cyprus", "Czechia", "Denmark", "Djibouti", "Dominica",
    "Dominican Republic", "Ecuador", "Egypt", "El Salvador", "Equatorial Guinea", "Eritrea", "Estonia", "Eswatini", "Ethiopia", "Fiji",
    "Finland", "France", "Gabon", "Gambia", "Georgia", "Germany", "Ghana", "Greece", "Grenada", "Guatemala",
    "Guinea", "Guinea-Bissau", "Guyana", "Haiti", "Honduras", "Hungary", "Iceland", "India", "Indonesia", "Iran",
    "Iraq", "Ireland", "Israel", "Italy", "Jamaica", "Japan", "Jordan", "Kazakhstan", "Kenya", "Korea, North",
    "Korea, South", "Kuwait", "Kyrgyzstan", "Laos", "Latvia", "Lebanon", "Lesotho", "Liberia", "Libya", "Liechtenstein",
    "Lithuania", "Luxembourg", "Madagascar", "Malawi", "Malaysia", "Maldives", "Mali", "Malta", "Mauritania", "Mauritius",
    "Mexico", "Moldova", "Monaco", "Mongolia", "Montenegro", "Morocco", "Mozambique", "Namibia", "Nepal", "Netherlands",
    "New Zealand", "Nicaragua", "Niger", "Nigeria", "North Macedonia", "Norway", "Oman", "Pakistan", "Panama", "Papua New Guinea",
    "Paraguay", "Peru", "Philippines", "Poland", "Portugal", "Qatar", "Romania", "Russia", "Rwanda", "Saint Kitts and Nevis",
    "Saint Lucia", "Saint Vincent and the Grenadines", "Samoa", "San Marino", "Sao Tome and Principe", "Saudi Arabia", "Senegal",
    "Serbia", "Seychelles", "Sierra Leone", "Singapore", "Slovakia", "Slovenia", "Solomon Islands", "Somalia", "South Africa",
    "South Sudan", "Spain", "Sri Lanka", "Sudan", "Suriname", "Sweden", "Switzerland", "Syria", "Taiwan*", "Tajikistan",
    "Tanzania", "Thailand", "Timor-Leste", "Togo", "Trinidad and Tobago", "Tunisia", "Turkey", "Uganda", "Ukraine", "United Arab Emirates",
    "United Kingdom", "Uruguay", "Uzbekistan", "Vanuatu", "Venezuela", "Vietnam", "West Bank and Gaza", "Yemen", "Zambia", "Zimbabwe"
  ),
  Population = c(
    40754388, 2832439, 44903225, 79824, 35588987, 93763, 45808747, 2780469, 26177413, 9042528,
    10358074, 409984, 1472233, 167885689, 281635, 9383862, 11655930, 400031, 13352864, 782455,
    12224110, 3233526, 2630296, 214326223, 459500, 6781953, 21497096, 54179306, 12889576, 593149,
    16767842, 27914536, 38250000, 5579144, 17179740, 19458310, 1411750000, 51516562, 836774, 5970424,
    89561403, 5180829, 26378274, 4030358, 11326616, 1207359, 10698896, 5882261, 988000, 72167,
    10953703, 17888474, 104258327, 6486205, 1402985, 3546421, 1326535, 1172362, 117876227, 896444,
    5540720, 67391582, 2225734, 2416668, 3989167, 83783942, 31072940, 10423054, 112523, 17915568,
    13132795, 1968001, 786552, 11402528, 10062991, 9660351, 364134, 1380004385, 273523615, 85028759,
    40462701, 5033165, 9291000, 60461826, 2961167, 125361201, 10203134, 19397998, 54985698, 25778816,
    51740000, 4270571, 6524195, 7275560, 1886198, 6825445, 2142249, 5180203, 6871292, 38128,
    2722289, 625978, 27691018, 19129952, 32365999, 521457, 20250833, 441543, 4649658, 1271768,
    128932753, 2617820, 39242, 3278290, 621718, 37457000, 31255435, 2540905, 29136808, 17134872,
    5122600, 6624554, 24206644, 206139589, 2083374, 5421241, 5106626, 220892340, 4314767, 8947024,
    7132538, 32971854, 109581078, 37846611, 10196709, 2881053, 19237691, 143447812, 13226796, 53199,
    183627, 110940, 198414, 33931, 219159, 35340683, 16743927, 8737371, 98347, 7976983,
    5850342, 5459642, 2078938, 686884, 15893222, 59308690, 11193725, 47615034, 22156000, 44909353,
    586634, 10099265, 8654622, 17500658, 23816775, 9537645, 59734218, 69950850, 1318445, 8278724,
    1399488, 11818619, 84680273, 45741007, 43733762, 9890402, 67886011, 3473730, 35163944, 307145,
    28435943, 97338579, 5250072, 30491000, 18383955, 14862924
  )
)

# Extract population data for US states from the us_deaths dataset and drop rows where Population is 0
population_data_states <- us_deaths %>%
  group_by(Province_State) %>%
  summarize(Population = sum(Population, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(Population > 0)  # Drop rows where Population is 0

# Add special cases with NA population (ships, etc.)
special_cases <- data.frame(
  Country.Region = c("Diamond Princess", "MS Zaandam", "Summer Olympics 2020", "Winter Olympics 2022", "Antarctica"),
  Population = NA
)

# Combine all population data for countries
population_data_complete <- rbind(
  population_data_countries,
  data.frame(Country.Region = "US", Population = sum(population_data_states$Population, na.rm = TRUE)),
  special_cases
)

global_analysis <- global_cases_long %>%
  filter(date == max(date)) %>%
  group_by(Country.Region) %>%
  summarize(total_cases = sum(cases)) %>%
  left_join(population_data_complete, by = "Country.Region") %>%
  mutate(
    cases_per_100k = (total_cases / Population) * 100000,
    cases_per_100k = round(cases_per_100k, 2)
  ) %>%
  arrange(desc(cases_per_100k)) %>%
  filter(!is.na(Population))

# Visualization for top 20 countries
ggplot(head(global_analysis, 20), 
       aes(x = reorder(Country.Region, cases_per_100k), y = cases_per_100k)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 20 Countries by COVID-19 Cases per 100,000 Population",
    x = "Country",
    y = "Cases per 100,000 Population"
  )

# US state analysis
us_analysis <- us_cases_long %>%
  filter(date == max(date)) %>%
  group_by(Province_State) %>%
  summarize(total_cases = sum(cases)) %>%
  left_join(population_data_states, by = "Province_State") %>%
  mutate(
    cases_per_100k = (total_cases / Population) * 100000,
    cases_per_100k = round(cases_per_100k, 2)
  ) %>%
  arrange(desc(cases_per_100k)) %>%
  filter(!is.na(Population))

# Visualization for top 20 US states
ggplot(head(us_analysis, 20), 
       aes(x = reorder(Province_State, cases_per_100k), y = cases_per_100k)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 20 US States by COVID-19 Cases per 100,000 Population",
    x = "State",
    y = "Cases per 100,000 Population"
  )
```

# Global cases map

```{r geographic_vis, include=TRUE}

# Prepare data for mapping
latest_date <- max(global_cases_long$date)
latest_data <- global_cases_long %>%
  inner_join(global_deaths_long, by = c("Country.Region", "Province.State", "Lat", "Long", "date")) %>%
  filter(date == latest_date) %>%
  group_by(Country.Region) %>%
  summarize(
    total_cases = sum(cases),
    total_deaths = sum(deaths),
    lat = mean(Lat, na.rm = TRUE),
    lng = mean(Long, na.rm = TRUE),
    cfr = (total_deaths/total_cases)*100
  ) %>%
  filter(!is.na(lat) & !is.na(lng))

# Create color palettes
pal_cases <- colorNumeric(
  palette = "viridis",
  domain = log10(latest_data$total_cases)
)

pal_cfr <- colorNumeric(
  palette = "RdYlBu",
  domain = latest_data$cfr,
  reverse = TRUE
)

map1 <- leaflet(latest_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lng,
    lat = ~lat,
    radius = ~sqrt(total_cases)/500,
    color = ~pal_cases(log10(total_cases)),
    fillOpacity = 0.7,
    popup = ~paste(
      "<strong>", Country.Region, "</strong><br/>",
      "Cases: ", formatC(total_cases, format="d", big.mark=","), "<br/>",
      "Deaths: ", formatC(total_deaths, format="d", big.mark=","), "<br/>",
      "CFR: ", round(cfr, 2), "%"
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_cases,
    values = ~log10(total_cases),
    title = "Log10(Total Cases)",
    opacity = 0.7
  )

# Case fatality rate map
map2 <- leaflet(latest_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lng,
    lat = ~lat,
    radius = 8,
    color = ~pal_cfr(cfr),
    fillOpacity = 0.7,
    popup = ~paste(
      "<strong>", Country.Region, "</strong><br/>",
      "CFR: ", round(cfr, 2), "%<br/>",
      "Deaths: ", formatC(total_deaths, format="d", big.mark=","), "<br/>",
      "Cases: ", formatC(total_cases, format="d", big.mark=",")
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_cfr,
    values = ~cfr,
    title = "Case Fatality Rate (%)",
    opacity = 0.7
  )

# Display maps
map1
map2

```

# Additional Visualizations

### Case fatality rate over time

```{r additional_vis, include=TRUE}

# Case fatality rate over time
global_data %>%
  mutate(fatality_rate = (total_deaths/total_cases)*100) %>%
  ggplot(aes(x=date, y=fatality_rate)) +
  geom_line(color="darkred", size=1) +
  labs(title="Global COVID-19 Case Fatality Rate Over Time",
       x="Date",
       y="Fatality Rate (%)") +
  theme_minimal()
```

### Wave Analysis - Daily New Cases Pattern

```{r additional_vis_wave, include=TRUE}

# Rolling Average Analysis
# Calculate daily new cases for top 5 countries
daily_cases_by_country <- global_cases_long %>%
  group_by(Country.Region, date) %>%
  summarize(cases = sum(cases)) %>%
  arrange(Country.Region, date) %>%
  group_by(Country.Region) %>%
  mutate(new_cases = cases - lag(cases, default = first(cases))) %>%
  ungroup()

# Wave Analysis - Daily New Cases Pattern
wave_analysis <- daily_cases_by_country %>%
  filter(Country.Region %in% top_5_countries) %>%
  mutate(month_year = floor_date(date, "month"))

# Wave Patterns
ggplot(wave_analysis, aes(x = date, y = new_cases, fill = Country.Region)) +
  geom_area(alpha = 0.5, position = "stack") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "COVID-19 Waves Over Time",
       subtitle = "Daily new cases for top 5 countries",
       x = "Date",
       y = "New Cases") +
  theme_minimal()

```

### Bubble Plot: Cases vs Deaths with Population Size

```{r additional_vis_bubble, include=TRUE}
# Bubble Plot: Cases vs Deaths with Population Size
# For top 20 countries at the latest date
latest_date <- max(global_cases_long$date)
bubble_data <- global_cases_long %>%
  inner_join(global_deaths_long, by = c("Country.Region", "date")) %>%
  filter(date == latest_date) %>%
  group_by(Country.Region) %>%
  summarize(total_cases = sum(cases),
            total_deaths = sum(deaths)) %>%
  arrange(desc(total_cases)) %>%
  head(20)

# Bubble Plot
ggplot(bubble_data, 
       aes(x = total_cases, y = total_deaths)) +
  geom_point(aes(size = total_cases), alpha = 0.6, color = "purple") +
  geom_text(aes(label = Country.Region), 
            check_overlap = TRUE, 
            vjust = -1, 
            size = 3) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  scale_size_continuous(range = c(3, 20)) +
  labs(title = "COVID-19 Cases vs Deaths",
       subtitle = "Size of bubble represents total cases",
       x = "Total Cases (log scale)",
       y = "Total Deaths (log scale)") +
  theme_minimal() +
  theme(legend.position = "none")
```

# Conclusions
Based on the comprehensive COVID-19 data analysis, several key findings emerge:

### Global Impact and Distribution
The data shows significant variation in COVID-19 impact across countries and regions
A strong correlation exists between total cases and deaths, though with varying case fatality rates (CFR)
Population-adjusted analysis reveals that some smaller countries were disproportionately affected when considering cases per 100,000 residents

### Temporal Patterns
Clear wave patterns are visible in the data, with multiple surges occurring at different times in different regions
The global CFR showed a general declining trend over time, possibly due to improved treatment protocols and vaccination efforts

### Geographic Variations
Significant regional differences in both case counts and mortality rates
Urban areas generally showed higher case counts, though this may partly reflect testing capacity
The mapping visualization reveals clusters of high-impact areas and potential geographic patterns in disease spread

# Potential Biases 

There could be inconsistencies between different countries and states in reporting

There could be inconsistent testing protocols across countries and regions

Potential underreporting in regions with limited healthcare infrastructure

Population density and age distribution are not fully accounted for

Healthcare system capacity are not factored in

Case fatality rates may be skewed by testing availability

Time lag between cases and deaths not fully addressed in the analysis

Vaccination rates are not included in the analysis
